---
title: "Assignment_Wk4_2026-10-04"
author: "Shivoan Balakumar"
date: "2026-09-29"
output: 
  html_document:
    theme: lumen
    highlight: pygments
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Assignment Week 4 - BIOS640

## Exercise 1

Install Packages
```{r install_pacman, echo = FALSE}
pacman::p_load(
  here,
  rio,
  tidyverse,
  viridis,
  dplyr,
  lubridate,
  readr,
  tydr,
  readxl,
  janitor,
  skimr,
  knitr,
  kableExtra,
  formatR,
  gridExtra,
  RColorBrewer,
  cowplot,
  ggpubr,
  ggokabeito,
  gghighlight,
  rcartocolor,
  DT
)
```

Import Data
```{r import_nhanes_data}
#import nhanes csv data set 
nhanes_data <- import(here("data", "cleaned_nhanes.csv"))

#provides dimensions - number of rows and columns
dim(nhanes_data)

# Shows the structure of data and data types
str(nhanes_data)

```

View Data
```{r view_data, eval = FALSE}
#provides dimensions - number of rows and columns
dim(nhanes_data)

# Shows the structure of data and data types
str(nhanes_data)
```
### Figures {.tabset}

#### Individual Figures

Plot 1 - Histogram of Age by Gender
```{r histogram_age_gender}
ggplot(data = nhanes_data, aes(x = age, fill = gender)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Age (years)") +
  theme(text = element_text(size = 12))
```

Plot 2 - Histogram of Ethnicity by Gender
```{r histogram_ethnicity_gender}
ggplot(data = nhanes_data, aes(x = ethnicity_2, fill = gender)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Ethnicity") +
  theme(text = element_text(size = 12))

```

#### Combined Figures

Combine Plots
```{r combine_plots}
#Create Plots
p1 <- ggplot(data = nhanes_data, aes(x = age, fill = gender)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Age (years)") +
  theme(text = element_text(size = 12))

p2 <- ggplot(data = nhanes_data, aes(x = ethnicity_2, fill = gender)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Ethnicity") +
  theme(text = element_text(size = 12))

# combine the plots with plot_grid
plot_grid(p1, p2,
           rel_widths = c(6, 8)
 )

# combine the plots with ggarrange
ggarrange(p1, p2, common.legend = TRUE, legend = "top")
  

```
Combining Plots with ggarrange looks better than combining with plot_grid because
it creates a common legend, leaving more width for the plots themselves.



## Exercise 2

### Explore the NHANES Dataset
```{r interactive_table}
DT::datatable(
  nhanes_data,
  options = list(scrollX = TRUE)
)
```


Plotting Age, Gender and Ethnicity_1 variables from NHANES
```{r plot_ethnicity_1}
plot_1 <- ggplot(nhanes_data, aes(x = age, fill = gender)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  facet_wrap(~ ethnicity_1) +
  labs(
    title = "Age Distribution by Gender and Ethnicity_1 Variable in NHANES Data",
    x = "Age (years)",
    y = "Number of participants")
plot_1
```
Plotting Age, Gender and Ethnicity_2 variables from NHANES
```{r plot_ethnicity_2}
plot_2 <- ggplot(nhanes_data, aes(x = age, fill = gender)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  facet_wrap(~ ethnicity_2) +
  labs(
    title = "Age Distribution by Gender and Ethnicity_2 Variable in NHANES Data",
    x = "Age (years)",
    y = "Number of participants")
plot_2
```


Showing distribution and missing Values of Age, Gender and Ethnicity variables
```{r}
summary(nhanes_data[c("age", "gender", "ethnicity_1", "ethnicity_2")])
```
The Ethnicity_2 Variable is more appropriate to keep because it provides more 
information and specificity on ethnicity in the study population 
(i.e., the "Asian" sub-population).

Saving the Two Plots
```{r}
ggsave(
  filename = here("figures", "plot1.png"),
  plot = plot_1,
  width = 8,
  height = 6
)

ggsave(
  filename = here("figures", "plot2.png"),
  plot = plot_2,
  width = 8,
  height = 6
)
```

## Exercise 3

The colleagues figure does not have a title, units for weight. It also doesn't 
easily demonstrate any "key messages" about the overall effectiveness of the diet.

This chart could be improved by measuring how much each participants weight changes
relatively compared to their baseline weight at week 0. That will provide more info 
on the diet's overall effect/participant weight change trends.

Creating New Columns for new plot demonstrating weight change from baseline at Week0
```{r new_data}
#importing data
diet_data <- import(here("data", "diet.csv"))

# Shows the structure of data 
str(diet_data)

#Creating new column for weight change from baseline at Week 0
diet_new <- diet_data %>%
  group_by(Participant) %>%
  mutate(
    baseline_weight = Weight[Week == 0],
    weight_change = Weight - baseline_weight
  )
# View new data
view(diet_new)
```

Creating new chart on diet data demonstrating weight change from baseline at Week0
```{r new_plot}
ggplot(diet_new, aes(
  x = Week,
  y = weight_change,
  colour = factor(Participant)
)) +
  geom_point() +
    labs(
    title = "New Diet Reduces Weight for Most Participants Over 16 Weeks",
    x = "Week",
    y = "Weight Change from Week 0",
    colour = "Participant"
  )
```


## Exercise 4

Exploring relationships between age, gender and SPB
```{r avg_SPB}
nhanes_data <- nhanes_data %>%
  mutate(
    avg_sbp = rowMeans(
      select(., systolic_bp_1, systolic_bp_2, systolic_bp_3, systolic_bp_4),
      na.rm = TRUE
    ),
    avg_sbp = ifelse(is.nan(avg_sbp), NA, avg_sbp)
  )

# Shows the structure of updated data 
str(nhanes_data)
```


